Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-12 draft.tex
Minor changes to title

\title{Insight Into Lumbar Back Pain  \subtitle{What the Lumbar Spine Tells About Your Life} }

\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
\affiliation{\sup{1}Department of Simulation and Graphics, University of Magdeburg}
\affiliation{\sup{2} Institut for Community Medicine, University of Greifswald}
\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald}
\email{\{paul, sylvia\}@isg.cs.uni-magdeburg.de, \{katrin.hegenscheid, voelzke\}@uni-greifswald.de}
}

\keywords{Epidemiology, Interactive Visual Analysis}

@@

Modern cohort studies often comprise medical image data.
%%
These data are hard to analyze.
%%
Individual diagnosis or manual segmentation of each body structure by radiologists is tedious, costly and comprises little reproducibility.
%%
Segmentation algorithms are not generally available and need to be custom-made for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
The lower (\emph{lumbar}) spine is the most stressed spine part.
%%
Lower back pain is one of the most frequent diseases in the western civilization.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, sporting activities and body posture combined with physical stress at work.
%%
The exact causes as well as particularly vulnerable risk groups are not known.
%%
Potential \emph{confounding} effects are also subject of current research.
%%

@@

%\com{Potentially skip these due to the confounding of body size}
\subsubsection{Body-Mass-Index Groups}
%%
Gender could be classified for each \emph{BMI} group using \emph{mean curvature} and \emph{curvature angle}.
%%
The error varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}, 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
%%
The starting age of smoking could be characterized well with an error between $0.25$ to $0.32$ for all \emph{BMI} groups except of subjects with BMI of \emph{$30.5-48$}.
%%
The result however is probably over fitted to the data due to tree sizes between $14$ and $16$.
%%
\com{Warum?}
%%

Some variables, such as \emph{body size} can be described with an error of $0.3$ to $0.36$ but only using large decision trees with over 20 rules.
%%
Notable is also the decreasing error with increasing decision tree size.
%%
Using mostly \emph{mean curvature} and \emph{curvature angle}, the leg pain level can be predicted using $14$ rules with an error of $0.46$ for obese subjects (\emph{BMI} higher than $30$).
%%
The dichotomous variable whether the subject has felt pain in the last seven days can also be predicted for this group using the same features.
%%
The resulting tree consists of $8$ rules and has an error of $0.35$.
%%
Obese subjects are prone to back and leg pain due to a more stressed lumbar spine.
%%
How much the spine is deformed under the heavy stress seems to directly influence the pain levels for these subjects.

@@

Subjects grouped using \underline{body size} have a similar spine shape and a lower variance.
%%
The remaining information within these groups is not enough to characterize back pain-related variables.
%%
For some \emph{body size} groups, certain \emph{nutrition} or \emph{psychological} parameters can be predicted with a comparatively high error.
%%

Another observation during the analysis is very large decision trees achieving low error scores.
%%
Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
%%
\subsection{Proposed Features}
\label{sec:ProposedFeatures}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.475\textwidth]{figures/proposed-features}
  \caption{
	Proposed features.
	%%
	(a) dented vertebraes are a sign of heavy stress and bone absorption.
	%%
	(b, c) the spine canal shape can be used to characterize scoliosis (curvature sagittal) and lordosis (curvature coronal) as well as the \emph{Cobb-Angles} \cite{Cobb}.
	%%
	(c) Spine canal thickness is associated with herniated disks.
	%%
	}
  \label{fig:proposedFeatures}
\end{figure}
%%
\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain \com{Referenz?}.
%%
We have to capture earlier signs of pathological change in order to derive better predictive models.
%%

>>>>

%\title{Insight Into Lumbar Back Pain  \subtitle{What the Lumbar Spine Tells About Your Life} }
\title{Interactive Visual Analysis of Lumbar Back Pain  \subtitle{What the Lumbar Spine Tells About Your Life} }

\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
\affiliation{\sup{1}Department of Simulation and Graphics, University of Magdeburg, Germany}
\affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
\email{\{klemm, glasser\}@isg.cs.uni-magdeburg.de}
}

\keywords{Epidemiology, Interactive Visual Analysis, Classification, Multi Modal Data}

@@

Modern cohort studies often comprise medical image data.
%%
These data are hard to analyze.
%%
Individual diagnosis or manual segmentation of each body structure by radiologists is tedious, costly and comprises little reproducibility.
%%
Segmentation algorithms are not generally available and need to be custom-made for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
%The lower (\emph{lumbar}) spine is the most stressed spine part .
%%
Lower back pain is one of the most frequent diseases in the western civilization \cite{Hoy2010}.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, sporting activities and body posture combined with physical stress at work.
%%
The exact causes as well as particularly vulnerable risk groups are not known.
%%
Potential \emph{confounding} effects are also subject of current research.
%%

@@

%\com{Potentially skip these due to the confounding of body size}
\subsubsection{Body-Mass-Index Groups}
%%
Gender could be classified for each \emph{BMI} group using \emph{mean curvature} and \emph{curvature angle}.
%%
The error varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}, 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
%%
The starting age of smoking could be characterized well with an error between $0.25$ to $0.32$ for all \emph{BMI} groups except of subjects with BMI of \emph{$30.5-48$}.
%%
The result however is probably over fitted to the data due to tree sizes between $14$ and $16$.
%%

Some variables, such as \emph{body size} can be described with an error of $0.3$ to $0.36$ but only using large decision trees with over 20 rules.
%%
Notable is also the decreasing error with increasing decision tree size.
%%
Using mostly \emph{mean curvature} and \emph{curvature angle}, the leg pain level can be predicted using $14$ rules with an error of $0.46$ for obese subjects (\emph{BMI} higher than $30$).
%%
The dichotomous variable whether the subject has felt pain in the last seven days can also be predicted for this group using the same features.
%%
The resulting tree consists of $8$ rules and has an error of $0.35$.
%%
Obese subjects are prone to back and leg pain due to a more stressed lumbar spine.
%%
The stress induced spine deformation seems to influence the pain levels for these subjects directly.
%How much the spine is deformed under the heavy stress seems to directly influence the pain levels for these subjects.

@@

Subjects grouped using \emph{body size} have a similar spine shape and a lower variance.
%%
The remaining information within these groups is not enough to characterize back pain-related variables.
%%
For some \emph{body size} groups, certain \emph{nutrition} or \emph{psychological} parameters can be predicted with a comparatively high error.
%%

Another observation during the analysis is very large decision trees achieving low error scores.
%%
Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
%%
\subsection{Proposed Features}
\label{sec:ProposedFeatures}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.475\textwidth]{figures/proposed-features}
  \caption{
	Proposed features.
	%%
	(a) dented vertebraes are a sign of heavy stress and bone absorption.
	%%
	(b, c) the spine canal shape can be used to characterize scoliosis (curvature sagittal) and lordosis (curvature coronal) as well as the \emph{Cobb-Angles} \cite{Cobb}.
	%%
	(c) Spine canal thickness is associated with herniated disks.
	%%
	}
  \label{fig:proposedFeatures}
\end{figure}
%%
\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain.
%%
We have to capture earlier signs of pathological change in order to derive better predictive models.
%%
